Comparison of Terrain Following and Cut Cell Grids using a Non-Hydrostatic Model

### 2015-12-11 src/tf-cutcell-comparison/responses2.tex
minor amendments for reviewer 2, round 2

\documentclass{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{natbib}
\usepackage{enumitem}
\usepackage{siunitx}
\usepackage{amsmath}
\usepackage{mathtools}
\newlist{comment}{enumerate}{1}
\setlist[comment,1]{label={\arabic{commenti}.},leftmargin=*,resume=comment}

@@

\section*{Review \#2}
\begin{quotation}
Overall, the authors have been very responsive to my comments and suggestions, and the manuscript improved considerably.
However, there are still a few minor aspects that need attention:
\begin{comment}
\item From equation (7a) it is clear that the symbol mu needs a physical unit which is 1/s. The statement in line 195
that mu is dimensionless is therefore wrong and needs to be corrected.
\end{comment}
\end{quotation}
\begin{quotation}
	\begin{comment}
	\item Line 241: $\phi$ is defined as the tracer density. Therefore, the definition of $\phi_0$ needs the physical unit \si{\kilogram\per\meter\cubed}
	\end{comment}
\end{quotation}
\begin{quotation}
	\begin{comment}
	\item Page 18, test c): The initial conditions look incomplete. Please provide information about the initial density or pressure initialization. Do you assume that the pressure at z=0 m is 1000 hPa?
	\end{comment}
\end{quotation}
\begin{quotation}
	\begin{comment}
	\item Test d) and e): As for test c), the descriptions of the initial conditions looks incomplete. Is the initial w zero in d,e?
\end{comment}
\end{quotation}
\begin{quotation}
	\begin{comment}
	\item Line 389: mu needs the physical unit 1/s.
\end{comment}
\end{quotation}
\begin{quotation}
	\begin{comment}
	\item Line 440: It would be better to say 'The functional form of the initial density field ...potential temperature field (Eq. (22)) in the gravity wave test.
	\end{comment}
\end{quotation}
\begin{quotation}
	\begin{comment}
	\item Line 445: While it is mathematically possible to initialize the tracer density with 288 (needs physical units!) at the ground, it does not make any physical sense. However, it is okay to leave it for this idealized test case, but a more physical choice is clearly more desirable.
	\end{comment}
\end{quotation}
\begin{quotation}
	\begin{comment}
	\item Line 449: Do you refer to the normalized $\ell_2$ error here?
	\end{comment}
\end{quotation}
\begin{quotation}
	\begin{comment}
	\item Check the references for some phrases that need to be capitalized, as e.g. Appalachian in line 571 or Lagrangian in line 567 
	\end{comment}
\end{quotation}

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{natbib}
\usepackage{enumitem}
\usepackage{siunitx}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{xcolor}
\newlist{comment}{enumerate}{1}
\setlist[comment,1]{label={\arabic{commenti}.},leftmargin=*,resume=comment}
\newcommand{\TODO}[1]{\textcolor{purple}{TODO: \emph{#1}}}

@@

\section*{Review \#2}
\begin{quotation}
Overall, the authors have been very responsive to my comments and suggestions, and the manuscript improved considerably.
However, there are still a few minor aspects that need attention:
\begin{comment}
\item From equation (7a) it is clear that the symbol mu needs a physical unit which is 1/s. The statement in line 195
that mu is dimensionless is therefore wrong and needs to be corrected.
\end{comment}
\end{quotation}
Removed `dimensionless' on line \TODO{xxx} and now included correct units where $\mu$ is specified on lines \TODO{xxx} and \TODO{xxx}.

\begin{quotation}
	\begin{comment}
	\item Line 241: $\phi$ is defined as the tracer density. Therefore, the definition of $\phi_0$ needs the physical unit \si{\kilogram\per\meter\cubed}
	\end{comment}
\end{quotation}
Added physical units to $\phi_0$ and in the caption for table 1.

\begin{quotation}
	\begin{comment}
	\item Page 18, test c): The initial conditions look incomplete. Please provide information about the initial density or pressure initialization. Do you assume that the pressure at z=0 m is 1000 hPa?
	\end{comment}
\end{quotation}
Yes, amended line \TODO{xxx} to state that $p_0 = \SI{1000}{\hecto\pascal}$, which applies to all tests that use the model.

\begin{quotation}
	\begin{comment}
	\item Test d) and e): As for test c), the descriptions of the initial conditions looks incomplete. Is the initial w zero in d,e?
\end{comment}
\end{quotation}
The value of $p_0$ applies to tests (c) and (d).  For test (d), line \TODO{xxx} now states that $w = \SI{0}{\meter\per\second}$.  For test (e), the velocity field is defined by the streamfunction in equation (26), and $w = \SI{0}{\meter\per\second}$ only where $\mathrm{d}h/\mathrm{d}x = 0$.

\begin{quotation}
	\begin{comment}
	\item Line 389: mu needs the physical unit 1/s.
\end{comment}
\end{quotation}
Added, many thanks.

\begin{quotation}
	\begin{comment}
	\item Line 440: It would be better to say 'The functional form of the initial density field ...potential temperature field (Eq. (22)) in the gravity wave test.
	\end{comment}
\end{quotation}
Instead of amending this line, we now refer to the field as a potential temperature field, not a tracer density field.  Further details are provided in response to the subsequent comment.

\begin{quotation}
	\begin{comment}
	\item Line 445: While it is mathematically possible to initialize the tracer density with 288 (needs physical units!) at the ground, it does not make any physical sense. However, it is okay to leave it for this idealized test case, but a more physical choice is clearly more desirable.
	\end{comment}
\end{quotation}
The field is really the potential temperature field from the gravity waves test.  We have modified this section so that the tracer density field is now called a potential temperature field.  Lines \TODO{xxx twice in the abstract, in the intro and conclusion} have also been amended accordingly so that this test is no longer refered to as a tracer advection test..  Equation (\TODO{xxx}) now has $\theta_0 = \SI{288}{\kelvin}$.

\begin{quotation}
	\begin{comment}
	\item Line 449: Do you refer to the normalized $\ell_2$ error here?
	\end{comment}
\end{quotation}
Yes, amended the manuscript to state this.
\begin{quotation}
	\begin{comment}
	\item Check the references for some phrases that need to be capitalized, as e.g. Appalachian in line 571 or Lagrangian in line 567 
	\end{comment}
\end{quotation}
Corrected these, also `Navier-Stokes' and `Met Office Unified Model'.

### 2015-12-11 src/tf-cutcell-comparison/tf-cutcell-comparison.tex
minor amendments for reviewer 2, round 2

\abstract{Terrain following coordinates are widely used in operational models but the cut cell method has been proposed as an alternative that can more accurately represent atmospheric dynamics over steep orography.  Because the type of grid is usually chosen during model implementation, it is typically necessary to use different models in order to compare the accuracy of different grids.  In contrast, here a single C-grid finite volume model is used to enable a like-for-like comparison of terrain following and cut cell grids.
A series of standard two-dimensional tests using idealised terrain are performed: tracer advection in a prescribed horizontal velocity field, a test starting from stably stratified initial conditions, and orographically induced gravity waves described by nonhydrostatic dynamics.  In addition, two new tracer advection tests are formulated having a velocity field that is everywhere tangential to the terrain following coordinate surfaces.  These new tests present a challenge to the advection scheme on cut cell grids.
The results of the advection tests demonstrate that tracer accuracy depends primarily upon alignment of the flow with the grid rather than the grid orthogonality.  As expected, the cut cell grid maintains greater accuracy for horizontal advection and in the test of stationary, stratified flow.  In the gravity waves test, results on all grids are in good agreement with existing results from the literature, although terrain following velocity fields lead to errors on cut cell grids.
Due to semi-implicit timestepping and an upwind-biased, explicit advection scheme, there were no timestep restrictions associated with small cut cells.}

@@

This study uses a modified version of the fully-compressible model from \citet{weller-shahrokhi2014} to enable a like-for-like comparison between terrain following and cut cell grids for idealised, two-dimensional test cases from the literature.  Section 2 presents the formulation of the terrain following and cut cell grids used in the experiments that follow.  In section 3 we give the governing equations, outline the model from \citet{weller-shahrokhi2014} and describe the modification which improves stability for long timesteps in the presence of steep orography.  Section 4 analyses the results from three tracer advection tests, a test of a stably stratified atmosphere initially at rest, and orographically induced gravity waves.  Concluding remarks are made in section 5.

@@

\subsection{Finite volume fully-compressible model}
\label{sec:euler-model}
The third model is taken from \citet{weller-shahrokhi2014} which details a discretisation of the fully-compressible Euler equations, given by
\begin{subequations}
\begin{align}
	\text{Momentum} &\ &\  	\frac{\partial \rho \mathbf{u}}{\partial t} + \nabla \cdot \rho \mathbf{u}\otimes\mathbf{u} &= \rho \mathbf{g} - c_p \rho \theta \nabla \Pi - \mu \rho \mathbf{u} \label{eq:momentum} \\
	\text{Continuity} &\ &\	\frac{\partial \rho}{\partial t} + \nabla \cdot \rho \mathbf{u} &= 0 \label{eq:cont} \\
	\text{Thermodynamic equation} &\ &\ \frac{\partial \rho \theta}{\partial t} + \nabla \cdot \rho \mathbf{u} \theta &= 0 \label{eq:theta} \\
	\text{Ideal gas law} &\ &\ \Pi^{(1 - \kappa)/\kappa} &= \frac{R \rho \theta}{p_0} \label{eq:state}
\end{align}
\end{subequations}
where \(\rho\) is the density, \(\mathbf{u}\) is the velocity field, \(\mathbf{g}\) is the gravitational acceleration, \(c_p\) is the heat capacity at constant pressure, \(\theta = T \left(p_0/p\right)^\kappa\) is the potential temperature, \(T\) is the temperature, \(p\) is the pressure, \(p_0\) is a reference pressure, \(\Pi = \left(p / p_0 \right)^\kappa\) is the Exner function of pressure, and \(\kappa = R/c_p\) is the gas constant to heat capacity ratio.  \(\mu\) is a dimensionless damping function used for the sponge layer in the gravity waves test in section~\ref{sec:results}\ref{sec:gw}.

@@

A series of two-dimensional tests are performed over idealised orography.  For each test, results on terrain following and cut cell grids are compared.  The first test from \citet{schaer2002} advects a tracer in a horizontal velocity field.  Second, a new tracer advection test is formulated employing a terrain following velocity field to challenge the advection scheme on cut cell grids.  The third test solves the Euler equations for a stably stratified atmosphere initially at rest, following \citet{klemp2011}.  Fourth, as specified by \citet{schaer2002}, a test of orographically-induced gravity waves is performed.  Finally, another tracer advection test is formulated that transports a stably stratified thermal profile in a terrain following velocity field.
No explicit diffusion is used in any of the tests.

@@

A tracer with density $\phi$ is positioned upstream above the height of the terrain.  It has the shape
\begin{align}
	\phi(x, z) &= \phi_0 \left\{ \begin{array}{l l}
		\cos^2 \left( \frac{\pi r}{2} \right) & \quad \text{if $r \leq 1$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.
%
\intertext{having radius, $r$, given by}
%
	r &= \sqrt{
		\left( \frac{x - x_0}{A_x} \right)^2 + 
		\left( \frac{z - z_0}{A_z} \right)^2
	}
\end{align}
where $A_x = \SI{25}{\kilo\meter}$, $A_z = \SI{3}{\kilo\meter}$ are the horizontal and vertical half-widths respectively, and $\phi_0 = 1$ is the maximum density of the tracer.  At $t = \SI{0}{\second}$, the tracer is centred at $(x_0, z_0) = (\SI{-50}{\kilo\meter}, \SI{9}{\kilo\meter})$ so that the tracer is upwind of the mountain and well above the maximum terrain height of \SI{3}{\kilo\meter}.  Analytic solutions can be found by setting the tracer centre such that $x_0 = ut$.
Tests are integrated forward in time for \SI{10000}{\second} with a timestep of \(\Delta t = \SI{25}{\second}\).

@@

The initial thermodynamic conditions are in discrete hydrostatic balance in the vertical direction, having a reference potential temperature of $\theta(z = 0) = \SI{288}{\kelvin}$ and constant stability with Brunt-V\"ais\"al\"a frequency $N = \SI{0.01}{\per\second}$ everywhere, except for a more stable layer of $N = \SI{0.02}{\per\second}$ between $\SI{2}{\kilo\meter} \leq z \leq \SI{3}{\kilo\meter}$.  The damping function, \(\mu\), is set to zero.  Unlike \citet{klemp2011}, there is no eddy diffusion in the equation set.

@@

A uniform horizontal wind $u = \SI{10}{\meter\per\second}$ is prescribed in the interior domain and at the inlet boundary.  No normal flow is imposed at the top and bottom boundaries and the velocity field has a zero gradient outlet boundary condition.

@@

Sponge layers are added to the upper \SI{10}{\kilo\meter} and leftmost \SI{10}{\kilo\meter} at the inlet boundary to damp the reflection of waves.
The damping function, \(\mu\), is adapted from \citet{melvin2010} such that
\begin{align}
	\mu(x, z) &= \mu_\mathrm{upper} + \mu_\mathrm{inlet} \\
	\mu_\mathrm{upper}(z) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{z - z_B}{H - z_B} \right) & \text{if } z \geq z_B \\
		0 & \text{otherwise} \\
	\end{cases} \\
	\mu_\mathrm{inlet}(x) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{x_I - x}{x_I - x_0} \right) & \text{if } x < x_I \\
		0 & \text{otherwise}
	\end{cases}
\end{align}
where $\overline{\mu} = 1.2$ is the damping coefficient, $z_B = \SI{20}{\kilo\meter}$ is the bottom of the sponge layer, $H = \SI{30}{\kilo\meter}$ is the top of the domain, $x_0 = \SI{-150}{\kilo\meter}$ is the leftmost limit of the domain and $x_I = \SI{-140}{\kilo\meter}$ is the rightmost extent of the inlet sponge layer.  The sponge layer is only active on faces whose normal is vertical so that it damps vertical momentum only.

@@

\subsection{Terrain following advection of thermal profile}
The potential temperature anomalies in the gravity waves test varied between terrain following and cut cell grids of the same resolution.  This variation may be due to differences in the wind fields between grids, or errors in the advection of potential temperature, amongst other possible causes.  To help establish the primary source of error, a new advection test is formulated in which the initial potential temperature field from the gravity waves test is used as a tracer density field.  To eliminate any differences in wind fields, the tracer is advected in a fixed, terrain-following velocity field that mimics the flow in the gravity waves test.

@@

The initial tracer density field, \(\phi\), and its boundary conditions, are the same as those of the initial potential temperature field in the gravity waves test.
Following the gravity waves test, the simulation is integrated forward by \SI{18000}{\second}, by which time the tracer initially upwind of the mountain will have cleared the mountain range.
Hence, the analytic solution, $\phi_T$, can be found by considering the vertical displacement of the tracer profile by the terrain following velocity field:
\begin{align}
	\phi_T(x, z) = \phi_0 \exp \left( \frac{N^2}{g} z^\star(x, z) \right) 
\end{align}
where the tracer density at $z = 0$, $\phi_0 = 288$, and the transform, $z^\star$, is given by rearranging equation~(\ref{eqn:btf}).

Enlargements of the error field near the mountain are shown in figure~\ref{fig:thermalAdvection} at $\Delta z = \SI{50}{\meter}$ with contours in tracer density overlayed.  Errors are too small to be visible on the BTF grid with a maximum error of \num{4.26e-4}.
However, on the cut cell grid, the maximum error is \num{0.302}.  Advection errors are apparent around mountainous terrain, with small cells having the largest errors.  These errors are advected horizontally along the lee slope, forming stripes.
The same error structure is present on all cut cell grids in which the terrain intersects the grid layers.

For comparison with the potential temperature anomalies in the gravity waves test, vertical profiles of tracer density error are taken at $x = \SI{50}{\kilo\meter}$.  As seen in figure~\ref{fig:gw-theta}c, errors are negligible on the BTF grids, but figure~\ref{fig:gw-theta}d reveals significant errors in the lowest layers of the cut cell grids that were advected from the mountain peaks.  The magnitude of error is about ten times larger than the potential temperature anomalies found on the cut cell grids in the gravity waves test seen in figure~\ref{fig:gw-theta}b.  We offer two explanations for this difference: first, there are no restoring forces in the linear advection model and, second, the wind fields in the two tests are different.

While the magnitude and structure of tracer error on the cut cell grids differs from potential temperature anomalies in the gravity waves test, results on the BTF grids are in close agreement in both tests but not on the cut cell grids.  Therefore, it is likely that anomalies on the cut cell grids in the gravity waves test are primarily due to errors in the advection of potential temperature through cut cells.

@@

Conversely, in the first terrain following tracer advection test, errors were large on the SLEVE and cut cell grids where velocities were misaligned with the grids.  Errors were also large on the cut cell grids in the second terrain following tracer advection test.  This result suggests that, in the gravity waves test, potential temperature errors in the cut cell grids are primarily due to advection errors.

@@

\begin{table*}
	\caption{Minimum and maximum tracer magnitudes and \(\ell_2\) error norms, defined by equation~(\ref{eqn:l2-error}), at \(t = \SI{10000}{\second}\) in the horizontal and terrain following tracer advection tests using centred linear and cubic upwind-biased schemes.  For the horizontal advection test, \(\ell_2\) error norms, minimum and maximum values are given for the fourth order scheme using the modified code from \citet{schaer2002}.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{l l l l S S S S}
\hline\hline
                  &                     &                  & Analytic & {BTF}   & {SLEVE}         & {Cut cell}     & {No terrain} \\
\hline
Horizontal        & Centred linear      & \(\ell_2\) error & 0        & 0.284   & 0.0316          & 0.0304         & 0.0304      \\
                  &                     & min              & 0        & -0.275  & -0.0252         & -0.0251        & -0.0251     \\
                  &                     & max              & 1        & 0.925   & 0.985           & 0.985          & 0.985       \\
                  & Fourth order        & \(\ell_2\) error & 0        & 0.0938  & 0.00244         & {---}          & 0.00234     \\
                  &                     & min              & 0        & -0.0926 & -0.00174        & {---}          & -0.00178    \\
                  &                     & max              & 1        & 1.00    & 0.984           & {---}          & 0.983       \\
                  & Cubic upwind-biased & \(\ell_2\) error & 0        & 0.112   & 0.0146          & 0.00784        & 0.00784     \\
                  &                     & min              & 0        & -0.0464 & -0.0106         & -0.000674      & -0.00674    \\
                  &                     & max              & 1        & 0.922   & 0.982           & 0.983          & 0.983       \\
\hline
Terrain following & Centred linear      & \(\ell_2\) error & 0        & 0.0341  & 0.235           & {unstable}     & {---}        \\
	          &                     & min              & 0        & -0.0245 & -0.120          & {unstable}     & {---}        \\
		  &                     & max              & 1        & 0.985   & 0.950           & {unstable}     & {---}        \\
		  & Cubic upwind-biased & \(\ell_2\) error & 0        & 0.0209  & 0.162           & 0.181          & {---}        \\
                  &                     & min              & 0        & -0.0110 & -0.0263         & -0.0284        & {---}        \\
                  &                     & max              & 1        & 0.983   & 0.865           & 0.851          & {---}        \\
\hline
\end{tabular}
\end{table*}

>>>>

\abstract{Terrain following coordinates are widely used in operational models but the cut cell method has been proposed as an alternative that can more accurately represent atmospheric dynamics over steep orography.  Because the type of grid is usually chosen during model implementation, it is typically necessary to use different models in order to compare the accuracy of different grids.  In contrast, here a single C-grid finite volume model is used to enable a like-for-like comparison of terrain following and cut cell grids.
A series of standard two-dimensional tests using idealised terrain are performed: tracer advection in a prescribed horizontal velocity field, a test starting from stably stratified initial conditions, and orographically induced gravity waves described by nonhydrostatic dynamics.  In addition, two new advection tests are formulated having a velocity field that is everywhere tangential to the terrain following coordinate surfaces.  These new tests present a challenge to the advection scheme on cut cell grids.
The results of the advection tests demonstrate that accuracy depends primarily upon alignment of the flow with the grid rather than the grid orthogonality.  As expected, the cut cell grid maintains greater accuracy for horizontal advection and in the test of stationary, stratified flow.  In the gravity waves test, results on all grids are in good agreement with existing results from the literature, although terrain following velocity fields lead to errors on cut cell grids.
Due to semi-implicit timestepping and an upwind-biased, explicit advection scheme, there were no timestep restrictions associated with small cut cells.}

@@

This study uses a modified version of the fully-compressible model from \citet{weller-shahrokhi2014} to enable a like-for-like comparison between terrain following and cut cell grids for idealised, two-dimensional test cases from the literature.  Section 2 presents the formulation of the terrain following and cut cell grids used in the experiments that follow.  In section 3 we give the governing equations, outline the model from \citet{weller-shahrokhi2014} and describe the modification which improves stability for long timesteps in the presence of steep orography.  Section 4 analyses the results from three advection tests, a test of a stably stratified atmosphere initially at rest, and orographically induced gravity waves.  Concluding remarks are made in section 5.

@@

\subsection{Finite volume fully-compressible model}
\label{sec:euler-model}
The third model is taken from \citet{weller-shahrokhi2014} which details a discretisation of the fully-compressible Euler equations, given by
\begin{subequations}
\begin{align}
	\text{Momentum} &\ &\  	\frac{\partial \rho \mathbf{u}}{\partial t} + \nabla \cdot \rho \mathbf{u}\otimes\mathbf{u} &= \rho \mathbf{g} - c_p \rho \theta \nabla \Pi - \mu \rho \mathbf{u} \label{eq:momentum} \\
	\text{Continuity} &\ &\	\frac{\partial \rho}{\partial t} + \nabla \cdot \rho \mathbf{u} &= 0 \label{eq:cont} \\
	\text{Thermodynamic equation} &\ &\ \frac{\partial \rho \theta}{\partial t} + \nabla \cdot \rho \mathbf{u} \theta &= 0 \label{eq:theta} \\
	\text{Ideal gas law} &\ &\ \Pi^{(1 - \kappa)/\kappa} &= \frac{R \rho \theta}{p_0} \label{eq:state}
\end{align}
\end{subequations}
where \(\rho\) is the density, \(\mathbf{u}\) is the velocity field, \(\mathbf{g}\) is the gravitational acceleration, \(c_p\) is the heat capacity at constant pressure, \(\theta = T \left(p_0/p\right)^\kappa\) is the potential temperature, \(T\) is the temperature, \(p\) is the pressure, \(p_0 = \SI{1000}{\hecto\pascal}\) is a reference pressure, \(\Pi = \left(p / p_0 \right)^\kappa\) is the Exner function of pressure, and \(\kappa = R/c_p\) is the gas constant to heat capacity ratio.  \(\mu\) is a damping function used for the sponge layer in the gravity waves test in section~\ref{sec:results}\ref{sec:gw}.

@@

A series of two-dimensional tests are performed over idealised orography.  For each test, results on terrain following and cut cell grids are compared.  The first test from \citet{schaer2002} advects a tracer in a horizontal velocity field.  Second, a new tracer advection test is formulated employing a terrain following velocity field to challenge the advection scheme on cut cell grids.  The third test solves the Euler equations for a stably stratified atmosphere initially at rest, following \citet{klemp2011}.  Fourth, as specified by \citet{schaer2002}, a test of orographically-induced gravity waves is performed.  Finally, another advection test is formulated that transports a stably stratified thermal profile in a terrain following velocity field.
No explicit diffusion is used in any of the tests.

@@

A tracer with density $\phi$ is positioned upstream above the height of the terrain.  It has the shape
\begin{align}
	\phi(x, z) &= \phi_0 \left\{ \begin{array}{l l}
		\cos^2 \left( \frac{\pi r}{2} \right) & \quad \text{if $r \leq 1$} \\
		0 & \quad \text{otherwise}
	\end{array} \right.
%
\intertext{having radius, $r$, given by}
%
	r &= \sqrt{
		\left( \frac{x - x_0}{A_x} \right)^2 + 
		\left( \frac{z - z_0}{A_z} \right)^2
	}
\end{align}
where $A_x = \SI{25}{\kilo\meter}$, $A_z = \SI{3}{\kilo\meter}$ are the horizontal and vertical half-widths respectively, and $\phi_0 = \SI{1}{\kilogram\per\meter\cubed}$ is the maximum density of the tracer.  At $t = \SI{0}{\second}$, the tracer is centred at $(x_0, z_0) = (\SI{-50}{\kilo\meter}, \SI{9}{\kilo\meter})$ so that the tracer is upwind of the mountain and well above the maximum terrain height of \SI{3}{\kilo\meter}.  Analytic solutions can be found by setting the tracer centre such that $x_0 = ut$.
Tests are integrated forward in time for \SI{10000}{\second} with a timestep of \(\Delta t = \SI{25}{\second}\).

@@

The initial thermodynamic conditions are in discrete hydrostatic balance in the vertical direction, having a reference potential temperature of $\theta(z = 0) = \SI{288}{\kelvin}$ and constant stability with Brunt-V\"ais\"al\"a frequency $N = \SI{0.01}{\per\second}$ everywhere, except for a more stable layer of $N = \SI{0.02}{\per\second}$ between $\SI{2}{\kilo\meter} \leq z \leq \SI{3}{\kilo\meter}$.  The damping function, \(\mu\), is set to \SI{0}{\per\second}.  Unlike \citet{klemp2011}, there is no eddy diffusion in the equation set.

@@

A uniform horizontal wind $(u, w) = (10, 0)\:\si{\meter\per\second}$ is prescribed in the interior domain and at the inlet boundary.  No normal flow is imposed at the top and bottom boundaries and the velocity field has a zero gradient outlet boundary condition.

@@

Sponge layers are added to the upper \SI{10}{\kilo\meter} and leftmost \SI{10}{\kilo\meter} at the inlet boundary to damp the reflection of waves.
The damping function, \(\mu\), is adapted from \citet{melvin2010} such that
\begin{align}
	\mu(x, z) &= \mu_\mathrm{upper} + \mu_\mathrm{inlet} \\
	\mu_\mathrm{upper}(z) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{z - z_B}{H - z_B} \right) & \text{if } z \geq z_B \\
		0 & \text{otherwise} \\
	\end{cases} \\
	\mu_\mathrm{inlet}(x) &= \begin{cases}
		\overline{\mu} \sin^2 \left( \frac{\pi}{2} \frac{x_I - x}{x_I - x_0} \right) & \text{if } x < x_I \\
		0 & \text{otherwise}
	\end{cases}
\end{align}
where $\overline{\mu} = \SI{1.2}{\per\second}$ is the damping coefficient, $z_B = \SI{20}{\kilo\meter}$ is the bottom of the sponge layer, $H = \SI{30}{\kilo\meter}$ is the top of the domain, $x_0 = \SI{-150}{\kilo\meter}$ is the leftmost limit of the domain and $x_I = \SI{-140}{\kilo\meter}$ is the rightmost extent of the inlet sponge layer.  The sponge layer is only active on faces whose normal is vertical so that it damps vertical momentum only.

@@

\subsection{Terrain following advection of thermal profile}
The potential temperature anomalies in the gravity waves test varied between terrain following and cut cell grids of the same resolution.  This variation may be due to differences in the wind fields between grids, or errors in the advection of potential temperature, amongst other possible causes.  To help establish the primary source of error, a new advection test is formulated in which the initial potential temperature field from the gravity waves test is used.  To eliminate any differences in wind fields, the field is advected in a fixed, terrain-following velocity field that mimics the flow in the gravity waves test.

@@

The potential temperature field, \(\theta\), and its boundary conditions, are the same as those of the initial potential temperature field in the gravity waves test.
Following the gravity waves test, the simulation is integrated forward by \SI{18000}{\second}, by which time the potential temperature initially upwind of the mountain will have cleared the mountain range.
Hence, the analytic solution, $\theta_T$, can be found by considering the vertical displacement of the thermal profile by the terrain following velocity field:
\begin{align}
	\theta_T(x, z) = \theta_0 \exp \left( \frac{N^2}{g} z^\star(x, z) \right) 
\end{align}
where the potential temperature at $z = 0$, $\theta_0 = \SI{288}{\kelvin}$, and the transform, $z^\star$, is given by rearranging equation~(\ref{eqn:btf}).

Enlargements of the error field near the mountain are shown in figure~\ref{fig:thermalAdvection} at $\Delta z = \SI{50}{\meter}$ with contours of potential temperature overlayed.  Errors are too small to be visible on the BTF grid with a maximum $\ell_2$ error of \num{4.26e-4}.
However, on the cut cell grid, the maximum error is \num{0.302}.  Advection errors are apparent around mountainous terrain, with small cells having the largest errors.  These errors are advected horizontally along the lee slope, forming stripes.
The same error structure is present on all cut cell grids in which the terrain intersects the grid layers.

For comparison with the potential temperature anomalies in the gravity waves test, vertical profiles of potential temperature error are taken at $x = \SI{50}{\kilo\meter}$.  As seen in figure~\ref{fig:gw-theta}c, errors are negligible on the BTF grids, but figure~\ref{fig:gw-theta}d reveals significant errors in the lowest layers of the cut cell grids that were advected from the mountain peaks.  The magnitude of error is about ten times larger than the potential temperature anomalies found on the cut cell grids in the gravity waves test seen in figure~\ref{fig:gw-theta}b.  We offer two explanations for this difference: first, there are no restoring forces in the linear advection model and, second, the wind fields in the two tests are different.

While the magnitude and structure of error on the cut cell grids in this test differs from potential temperature anomalies in the gravity waves test, results on the BTF grids are in close agreement in both tests but not on the cut cell grids.  Therefore, it is likely that anomalies on the cut cell grids in the gravity waves test are primarily due to errors in the advection of potential temperature through cut cells.

@@

Conversely, in the first terrain following tracer advection test, errors were large on the SLEVE and cut cell grids where velocities were misaligned with the grids.  Errors were also large on the cut cell grids in the second terrain following thermal advection test.  This result suggests that, in the gravity waves test, potential temperature errors in the cut cell grids are primarily due to advection errors.

@@

\begin{table*}
	\caption{Minimum and maximum tracer densities (\si{\kilogram\per\meter\cubed}) and \(\ell_2\) error norms, defined by equation~(\ref{eqn:l2-error}), at \(t = \SI{10000}{\second}\) in the horizontal and terrain following tracer advection tests using centred linear and cubic upwind-biased schemes.  For the horizontal advection test, \(\ell_2\) error norms, minimum and maximum values are given for the fourth order scheme using the modified code from \citet{schaer2002}.}
\label{tab:advection}
%
\centering
\footnotesize
\begin{tabular}{l l l l S S S S}
\hline\hline
                  &                     &                  & Analytic & {BTF}   & {SLEVE}         & {Cut cell}     & {No terrain} \\
\hline
Horizontal        & Centred linear      & \(\ell_2\) error & 0        & 0.284   & 0.0316          & 0.0304         & 0.0304      \\
                  &                     & min              & 0        & -0.275  & -0.0252         & -0.0251        & -0.0251     \\
                  &                     & max              & 1        & 0.925   & 0.985           & 0.985          & 0.985       \\
                  & Fourth order        & \(\ell_2\) error & 0        & 0.0938  & 0.00244         & {---}          & 0.00234     \\
                  &                     & min              & 0        & -0.0926 & -0.00174        & {---}          & -0.00178    \\
                  &                     & max              & 1        & 1.00    & 0.984           & {---}          & 0.983       \\
                  & Cubic upwind-biased & \(\ell_2\) error & 0        & 0.112   & 0.0146          & 0.00784        & 0.00784     \\
                  &                     & min              & 0        & -0.0464 & -0.0106         & -0.000674      & -0.00674    \\
                  &                     & max              & 1        & 0.922   & 0.982           & 0.983          & 0.983       \\
\hline
Terrain following & Centred linear      & \(\ell_2\) error & 0        & 0.0341  & 0.235           & {unstable}     & {---}        \\
	          &                     & min              & 0        & -0.0245 & -0.120          & {unstable}     & {---}        \\
		  &                     & max              & 1        & 0.985   & 0.950           & {unstable}     & {---}        \\
		  & Cubic upwind-biased & \(\ell_2\) error & 0        & 0.0209  & 0.162           & 0.181          & {---}        \\
                  &                     & min              & 0        & -0.0110 & -0.0263         & -0.0284        & {---}        \\
                  &                     & max              & 1        & 0.983   & 0.865           & 0.851          & {---}        \\
\hline
\end{tabular}
\end{table*}
